\section{Introduction}
%Write around four paragraphs establishing the context and motivating your project.
The aim of this project is to design a heterogeneous system on a chip (SoC) using the RISC-V open standard ISA\cite{riscv-1}\cite{riscv-2}. Once designed, the SoC should be implementable on an FPGA and be able to execute bare-metal C code and assembly with SMP, allowing all cores to execute code at the same time.

\subsection{Motivation}
Designing a processor is a complex challenge. They need to be power efficient to reduce the cost of running the system, as well as providing adequate processing power when required. Often one is sacrificed to improve the other: for instance, energy efficiency is often reduced due in designs of large processors containing multiple cores. Due to the increased size and complexity, more power has to be supplied to achieve the same performance as a smaller processor.

A heterogeneous CPU attempts to solve this issue by combining dissimilar core designs; often a powerful core, or p-core, and an efficient core, or e-core. When only light processing is required, the p-core can be effectively shutdown and the e-core will do all processing, resulting in less power used. For heavy processing, the p-core is then used to increase peak performance. Depending on the exact implementation, the p-core can be used either individually or in tandem with the e-core, but both result in greater performance than just the e-core.

This has been used extensively in many mobile devices, and increasingly in larger devices like laptops. The Apple M2 processor\cite{applem2} is one such example, implementing 4 high-performance cores and 4 energy-efficient cores. These are arranged in a hybrid configuration similar to ARM DynamIQ and big.LITTLE\cite{biglittle}, allowing cores to dynamically be assigned work that best fits them. In Apple's design, both cores are capable of executing the same code, with the performance core benefitting from much increased cache and other proprietary changes to increase it's speed over the efficiency core. 

\subsection{Related work}
%TODO: To be updated - mostly copied from specification, want this to be much more critical/analytical
%Discuss related work.
\subsubsection{A RISC-V Heterogeneous SoC for Embedded Devices\cite{valenterisc}}
This project is ongoing, and presents work designing a RV64 (RISC-V 64-bit) host core that offloads tasks to a PMCA (Programmable Many Core Accelerator) made from RV32 (RISC-V 32-bit) cores, which implement extensions for machine learning and discrete signal processing. The suggested use-case for the SoC is in IoT applications and programmable embedded devices. The host core is Linux compatible, and offers a full OS that acts as a platform for programs that run on the PMCA. The use of a large RV64 core to allow a full Linux OS to run on the SoC provides a huge amount of flexibility to the programmer, as the OS implements features like CLI, memory virtualisation, networking and more that allow programs to be written much more generally than embedded software running without an OS. However, this usage of a full Linux OS could be considered excessive for the use-case. An embedded Linux OS would have a lower overhead due to the reduced services it offers, which is very beneficial in an embedded environment where efficiency is highly important. Unfortunately, no data is provided about the processing power or energy usage of the design.

\subsubsection{Muntjac multicore RV64 processor\cite{UCAM-CL-TR-972}}
Muntjac is an SoC generator comprising of multiple components, that can be used to produce a Linux capable SoC. There is only one type of core in the system, RV64, but the SoC can be multicore. The project report is dedicated to the design of the core and cache as opposed to usage in any devices, as the purpose is to provide an easily understood and extensible platform for specialised designs. This is excellently done - the project is very well presented and uses multiple open-standards like TileLink\cite{tilelink} to increase the ease of working with it. It would be possible for this project to be extended for a heterogeneous SoC design, but it appears that this has not yet been done in any public works that extend the project.

\subsection{Objectives}
%One sentence summary of your project. Followed by a short list of concrete objectives:
The overall aim is design a heterogeneous SoC containing two types of RISC-V core that can execute bare-metal C code and assembly with SMP. Objectives have been labelled according to the MoSCoW method\cite{case-method-fasttrack} to indicate their importance and expected completion.
\begin{figure}[h!]
    \centering
    \begin{enumerate}
        \item Design a heterogeneous RISC-V SoC containing 2 dissimilar cores, each capable of executing at least the RV64 instruction set (Must).
        \item Execute assembly instructions on both cores when implemented in an SoC on an FPGA (Must).
        \item Execute bare-metal code on both cores when implemented in an SoC on an FPGA (Should).
        \item Measure the performance of the SoC for comparison against single-core designs. (Should)
        \item Run embedded Linux on the SoC and connect to it via serial or SSH (Could).
        \item Allow processes to execute on both cores inside of embedded Linux (Could).
        \item Intelligently select which core the process will run on, depending on factors such as process priority and resource usage (Won't).
    \end{enumerate}
    \caption{Objectives for the project}
    \label{fig:objectives}
\end{figure}